---
title: "User Based Collaborative Filtering"
author: "Audrey Pentz"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{User Based Collaborative Filtering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  echo = TRUE,
  comment = "#>"
)

# load packages
library(cf)
library(dplyr)
library(tidyr)


```


### Objectives

User-Based Collaborative Filtering is 1 of 3 techniques used to build a Recommender System in the cf package. A demonstration of how to use the cf package for User-Based Collaborative Filtering follows. The demonstration uses a book ratings dataset of 10,000 users that have rated 150 books on a scale from 0 to 10 and is loaded with the package. The package can be used in this way on any similar dataset. 

The objectives of this vignette are as follows:  
\n- To transform the data into the format required  
\n- To calculate cosine similarity scores between Users (rows)  
\n- To make a recommendation for 1 or more existing Users  
\n- To make a recommendation for a new User  


### 1. Transform the data


#### 1.1. Remove implicit (zero) ratings and NAs

We are only going to use the \code{book_ratings} data. Ratings are on a scale from 0 to 10 but not all readers have rated all books. Zero ratings are called implicit ratings where the reader has read the book but not rated it. A rating from 1 to 10 is called an explicit rating where the reader has actively chosen a rating for the book they have read. So we need to remove the implicit (zero ratings). NAs must also be removed, although this is not applicable to this dataset. The function that does this is called \code{remove_zero_blanks} and can be applied as shown.

```{r remove}

data <- cf::remove_zeros_blanks(book_ratings)
head(data, 10)

```


#### 1.2. Remove Duplicates

It would make more sense to see the book names rather than ISBNs. In the Exploratory Data Analysis Vignette (EDA) we saw that 2 readers have rated the same book twice. Reader `11676` has given 2 differet ratings for the same book (`the nanny diaries: a novel`), 6 and 9. Reader `251140` has given the same rating for the same book (`the nanny diaries: a novel`) twice, both rated 10. Therefore, converting from long to wide will fail unless we remove these duplicates from the data first, as shown below.

```{r dedupe}

library(dplyr)

# identify the duplicates
book_ratings_with_names <- data %>% 
  left_join(book_info) %>%
  select(User.ID, ISBN, Book.Title, Book.Rating) %>% 
  filter(Book.Rating %in% c(1:10))
book_ratings_with_names[c("1008","1009","16884","16885"),]

# let's get the average rating for book 'the nanny diaries: a novel'
book_ratings %>% 
  left_join(book_info) %>%
  select(Book.Title, Book.Rating) %>% 
  filter(Book.Title == 'the nanny diaries: a novel') %>% 
  filter(Book.Rating %in% c(1:10)) %>% 
  summarise(avg_bookrating = mean(Book.Rating))

# let's get the average rating for reader 11676
book_ratings %>% 
  filter(User.ID == '11676') %>% 
  summarise(avg_readerrating = mean(Book.Rating))
# the mean rating for this book is 7.35 which is closer to 6 than to 9
# the mean rating for this user is 5.05 which is closer to 6 than to 9
# so we will discard the second rating for reader 11676
# reader 251140 gave the same rating so it doesn't matter which is removed

duplicates <- book_ratings_with_names[c("1009","16885"),]
duplicates
book_ratings_with_names_deduped <- book_ratings_with_names %>% 
  anti_join(duplicates) %>% 
  select(User.ID, Book.Title, Book.Rating)


```



#### 1.3. Transform from long to wide

We now need to reshape the data so that we have a matrix with `Users` (readers in this case) in the rows and `Items` (books in this case) in the columns, with `ratings` in the body of the matrix. This is demonstrated with the \code{long_to_wide} function.

```{r wide}

data <- cf::long_to_wide(book_ratings_with_names_deduped, fill = 0)
data[c(8,30,35,39,40),1:10]

```


#### 1.4. Convert to matrix and remove column 1 (UserID)

We now need to reshape the data so that we have a matrix with `Users` (readers in this case) as the rownames and `Items` (books in this case) as the column names, with `ratings` in the body of the matrix. This is demonstrated with the \code{convert_to_matrix} function.

```{r matrix}

data <- cf::convert_to_matrix(data)

# let's take a look at ratings for a for a selection of readers and books
demo <- data[c("23872","63360","98783","16795", "11676"),
     c("1st to die: a novel","angela's ashes", "black and blue",
       "fahrenheit 451", "hannibal", "silence of the lambs",
       "interview with the vampire", "stones from the river", 
       "the fellowship of the ring (the lord of the rings, part 1)",
       "the horse whisperer","the vampire lestat (vampire chronicles, book ii)")]
# let's transpose it to make it easier to display
t(demo)

```

Just from these ratings we can already form the impression that reader `16795` and `11676` have given some books similar ratings eg. `1st to die: a novel` and `angela's ashes` and some books different ratings such as `the horse whisperer`. Reader `23872` and `63360` have also given some books similar ratings such as `interview with the vampire` and `the fellowship of the ring (the lord of the rings, part 1)`. Reader `11676` differs with reader `23872`'s rating of `interview with the vampire` by 3 and with their rating of `the vampire lestat (vampire chronicles, book ii)` by 2.


### 2. Calculate Cosine Similarity Scores between Users

The cf package uses cosine similarity to calculate similarity scores between pairs of Users (readers in this case). This is done using the \code{user_similarities} function.

```{r similarities}

# calculate cosine similarity scores for all users
user_sims <- cf::user_similarities(data)

# let's look at the scores for a few users we now know
user_sims[c("23872","63360","98783","16795", "11676"),
          c("23872","63360","98783","16795", "11676")]


```

From the ratings, we formed the impression that reader `16795` and `11676` are similar and their cosine similarity is 0.5038 and reader `23872` and `63360` are similar and their cosine similarity is 0.7070 which makes sense.



We can also have a look at which readers are most similar to a given reader, like this:

```{r similar}

# who are the most similar reader to a given reader?
head(sort(user_sims["23872",], decreasing = TRUE),8)
head(sort(user_sims["63360",], decreasing = TRUE),8)
head(sort(user_sims["98783",], decreasing = TRUE),8)
head(sort(user_sims["16795",], decreasing = TRUE),8)
head(sort(user_sims["11676",], decreasing = TRUE),8)

```


### 3. Make a prediction for 1 or more Users

Now that we have the similarity scores for every pair of Users (readers), we can see which Users are similar to other Users in the dataset.

```{r predict}

user_predictions <- user_predict(user = "23872",
                                  user_similarities = user_sims,
                                  ratings_wide = data, 
                                  predict_n = 10)
user_predictions

```

Let's look at recommendations for all of the demo users chosen above

```{r predict_demo}

lapply(c("23872","63360","98783","16795", "11676"), user_predict, user_sims, data, 10)

```


### 4. Make a prediction for a new User

```{r predict_new}

# create data for new users
users_new <- data.frame(User.ID = (rep(max(book_ratings_with_names_deduped$User.ID)+1,3)), 
                        ISBN = c("0440234743", "0971880107", "0345417623"), 
                        Book.Rating = c(2, 5, 3))
# get the User.ID of the new user #"278844"
max(book_ratings_with_names_deduped$User.ID)+1

# remove ISBN and add Book.Title 
users_new <- users_new %>% 
  left_join(book_info, by = "ISBN") %>% 
  select(User.ID,Book.Title,Book.Rating)

# attach the new user to the data
book_ratings_new <- rbind(book_ratings_with_names_deduped, users_new)

# reshape the data from long to wide
book_ratings_new_wide <- cf::long_to_wide(book_ratings_new, fill = 0)

# convert to matrix
book_ratings_new_wide <- cf::convert_to_matrix(book_ratings_new_wide)

# recalculate similarities
user_sims_new <- cf::user_similarities(book_ratings_new_wide)

# generate predictions
user_predictions <- user_predict(user = "278844",
                                  user_similarities = user_sims_new,
                                  ratings_wide = book_ratings_new_wide, 
                                  predict_n = 10)
user_predictions


```


